Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-06 Manuscript/manuscript.tex

In this work, we empirically sample this parameter space $f_q \in [0.75, 1.0]$
to find an atomistic MD simulation force field parameters for lipids that
accurately describe the lipid head group response to varying concentrations of cations
when compared to NMR data \cite{akutsu81,altenbach84,scherer89}. Such data can be used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the headgroup,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and headgroup
parameters are required in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, applied to headgroup, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related headgroup order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of headgroup, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Comparison of ion binding affinity to experiments by using the electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in headgroup, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of headgroup order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

\subsection{Ion concentrations and binding affinity}
The experimental headgroup order parameter data used
in this work is reported as a function of 
cation concentration remaining in the bulk 
solvent \cite{akutsu81,altenbach84}. 
To determine the same concentration from simulations, we
use large enough box to ensure that the ion concetration
reaches a clear plateau and the electroneutrality
is satisfied at the edge of simulation box. It should be noted that the recent work
by NMRlipids project compared simulations to the same 
experimental data by using the relative molar concentration in water
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M \cite{catte16}.
This is not exactly the same as the concentration reported in the used experimental
data, because the binding of ions in bilayer reduces the bulk concentration.
The difference slightly moves points towards lower concentrations but do not 
affect the conclusions in the previous study \cite{catte16}.

The ion binding affinities between different models were compared by calculating
the relative surface excess of ions with respect to water, $\Gamma_i^w$, as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that phospholipid membranes are bilayers, 
and hence the interface occurs in the simulation box twice
providing a surface area $2A$.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, can be determined
from the concentration plateau at the edge of the simulation box.

\subsection{Comparison of lipid bilayer structure to experiments}

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
In particular, the periodic orthorhombic simulation box conatined 
128 POPC molecules and approximately 50~water molecules per each lipid.
Water molecules were described by OPC3 model~\cite{Izadi16} 
as it is currently the most accurate three site rigid water model. 
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}. 
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
The systems were simulated for 200~ns using Lipid14 model for POPC.
First 20~ns was omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC \todo{How long?}.
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all \todo{Were all scaled? Are there charges is tails? Were those scaled as well in this case?} 
charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{There seems to a typo in experimental value for other order parameter for C2 carbon in sn-2 chain; it should be negative.
  The scale of the y-axis should be then zoomed when this is fixed.} \\
  \todo{x-axis scale in form factor figure from 0 to 0.6, until where experimental data is available.}
\end{figure}

@@

%depends on both, the
%binding affinity and the response of the headgroup on bound charge.
%Thus, the headgroup response to bound charge has to be correctly
%described in the model if the concept is used for quantitative studies of
%binding affinity. Hence, we first quantify the response of head group order
%parameters to the bound charge by using experimental data measured from
%mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
%
%Thus, the headgroup order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The headgroup order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the headgroup order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Changes of headgroup order parameters of POPC bilayer as a function of NaCl and CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the plateau of cation number density $C_{np}$
    at the box edge as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure*}

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 nm^{-3}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = XX nm^{-3}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 nm^{-3}$ with Dang ions.
\todo{Surface excess for Lipid14 with Aqvist yet to be added.} \\
\todo{Surface excess units are not correct. Maybe the numbers needs to be checked also?} \\
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.

>>>>

In this work, we empirically sample this parameter space $f_q \in [0.75, 1.0]$
to find an atomistic MD simulation force field parameters for lipids that
accurately describe the lipid head group response to varying concentrations of cations
when compared to NMR data \cite{akutsu81,altenbach84,scherer89}. Such data can be used to accurately asses
ion binding in PC bilayers, as discussed in Ref, \citenum{catte16} and in section \ref{section:electrometer}.
We chose Lipid14~\cite{dickson14} force field parameters as a starting point,
because their response to bound ions was apparently most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding.
The hydrocarbon chain parameters are not modified, because they are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in various conditions
in most lipids models, including Lipid14 \cite{ollila16}.
In contrast, improvements in glycerol backbone and head group
parameters are required in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, applied to head group, glycerol backbone
and carbonyl charges of Lipid14 model, we found out that ion binding and
related head group order parameter responses become weaker. 
The optimal behaviour of ion binding was observed with
the scaling factor $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is significantly smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of area was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stonger electrostatic interactions with them.
After reducing the $\sigma$ parameters with a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
again in agreement with experimental value (see Table~\ref{tab:apls}).

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding was compared between experiments and simulations by 
using lipid head group order parameters and 
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of bound charge in a lipid bilayer \cite{seelig87}.
This is rationalized as a change of lipid head group dipole tilt to more vertical orientation
with bound positive charge and {\it vice versa} for negative charge \cite{seelig87}. 
More recent analysis in the NMRlipids project concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. The concept was also shown
to be suitable for the comparison of ion binding affinity in lipid 
bilayers between simulations and experiments~\cite{catte16}.

The order parameters for $\alpha$ and $\beta$ segments in head group, as well as for
all C-H bonds in lipid molecules,
can be accurately measured by using $^2$H NMR or $^{13}$C NMR techniques.
The order parameters can be calculated from MD simulations 
by using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and average is taken over all sampled configurations \cite{ollila16}.

According to the electrometer concept,
the change of the head group order parameters as a function of
the amount of bound charge per lipid $X^\pm$ 
can be written as~\cite{ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) =\frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
%the empirically observed linear relation %for the change in order parameters %with respect to a bilayer without bound charges              
%\begin{equation}\label{electrometer_eq}                                                                                                   
%S_{\rm{CH}}^{i}(X^\pm)=S_{\rm{CH}}^{i}(0) + \frac{4m_i}{3\chi} X^\pm,                                                                     
%\end{equation}                                                                                                                            
$m_i$ is an empirical constant depending on the valency and position of bound charge,
and the value of the quadrupole coupling constant is $\chi \approx$\,167\,kHz.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$  and $m_\beta=-10.0$ for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
Before using electrometer concept to analyze cation binding, 
we calibrate the response of head group order parameters to the bound
charge in the simulation models. This is done by comparing 
the slope $m_i$ in Eq. \ref{OPchangeEQ} between simulations and
experiments of cationic surfactants in POPC bilayer \cite{scherer89}. 
Since essentially all surfactants locate in bilayer, the 
molar ratio directly gives the amount of bound charge per 
lipid $X^\pm$ is these systems.

\subsection{Ion concentrations and binding affinity}
The salt concentrations are reported in two different ways
in the used experimental data for the lipid head group order
parameters. The data for DPPC is reported
by using solution concentrations before solvating
the lipids~\cite{akutsu81}, while the data
for POPC~\cite{altenbach84} is reported by using
supernatant concetration measured with atomic absorption
spectroscopy after solvation of lipids.
The former was used in comparison with experiments and
calculated from simulations in previous work as 
[salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M~\cite{catte16}.
Here we use the latter, because the focus is on POPC bilayers.
This is determined from the ion concentrations, which reach
a constant value at the edges of large enough simulation box.
The two concentrations differ somewhat for CaCl$_2$ systems, because some
amount of ions bind in lipids. However, the small difference do
not affect the conclusions in this or in previous work~\cite{catte16}.

The ion concentrations with constant value at the simulation box edges
are also used to quantify the ion binding affinities is different simulation
models by using the relative surface excess of ions with respect to water $\Gamma_i^w$.
This is defined as \cite{chattorajBOOK}
\begin{eqnarray}
  \Gamma_i^w   &=& \Gamma_i-\Gamma_w \frac{C_i-C_i'}{C_w-C_w'} \\
  \Gamma_{a} &=& \frac{n_{a} - C_{a} V - C_{a}' V'}{A} \, \, ; \, a=i,w \, ,
\end{eqnarray}
where index $i$ and $w$ denote ions and water respectively,
prime accent denotes the phase inside bilayer, 
whereas non-prime symbols account for bulk water phase,
$n$ is the total amount of water or ions,
$C$ denotes bulk concentrations, 
$V$ is the volume of the respective phase separated by the Gibbs divinding plane, 
$A$ is the area of the interface, 
and $\Gamma_a$ is the surface excess of water or ions. 
As $\Gamma_i^w$ does not depend on the definition of the interfacial region
and the position of the Gibbs dividing plane~\cite{chattorajBOOK},
we consider the whole simulation box as an interface. 
Note that the area of the interface is given by the simulation box area
multiplied with two $2A_b$, because bilayers have a interface
in both leaflefts.
The concentrations of water and ions inside
the bilayer can be expected to vanish, thus $c_{i,w}'=0$.
This provides us a simplified relation for 
the relative surface excess of lipid bilayers in simulations
\begin{equation}\label{surfexcess}
  \Gamma_i^w=\frac{1}{2A_b} \left ( n_i - n_w \frac{C_i}{C_w} \right ) ,
\end{equation}
where the bulk concentrations, $C_w$ and $C_i$, are 
concentrations with constant value at the simulation box edges.

\subsection{Validation of lipid bilayer structure against experiments}

@@

\subsubsection{Simulations with aqueous ions}
The simulated systems consisted of 1-Palmitoyl-2-oleoylphosphatidylcholine
(POPC) bilayer and an aqueous salt solutions of varying concentrations.
In particular, the periodic orthorhombic simulation box conatined 
128 POPC molecules and approximately 50~water molecules per each lipid.
Water molecules were described by OPC3 model~\cite{Izadi16} 
as it is currently the most accurate three site rigid water model. 
In order to test transferability of the newly developed ECC-lipids model, 
we also performed several additional simulations with water models 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005}
presented in Supporting Information (SI). 
We used ECC-ions model for ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulations with Lipid14 use ion models by Dang~\cite{smith94,chang1999,dang2006}, and by \AA{}qvist~\cite{aqvist90}. 
Classical molecular dynamics simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}, 
and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
  this is done in Amber by default. If this is correct, the information in the
  table do not match with this sentence. Based on semi-extensive testing I made few years
ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

\subsubsection{Simulations with cationic surfactants}
Automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium bromide, C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$, molecule.
AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions 10\%, 20\%, 30\%, 42\% and 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35 or 50 surfactants per leaflet.
The systems were simulated for 200~ns using Lipid14 model for POPC.
First 20~ns were omitted from the analysis.

The same systems were also simulated with ECC-lipid model for POPC using the same setup. 
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.0cm]{../Fig/POPCstructure.eps}
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.eps}
  \includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    X-ray scattering form factors from experiments~\cite{Kucerka2011} and simulations using Lipid14 \cite{dickson14} and ECC-lipids models. 
    Order parameters of head group, glycerol backbone and sn-1 and sn-2 tails  from simulations with Lipid14 \cite{dickson14} and ECC-lipids models
    compared with experimental order parameters from \cite{ferreira13}.}
  \todo{There seems to a typo in experimental value for other order parameter for C2 carbon in sn-2 chain; it should be negative.
    The scale of the y-axis should be then zoomed when this is fixed.} \\
  \todo{I would put the acyl chain order parameters also for original Lipid14}\\
  \todo{x-axis scale in form factor figure would be better from 0 to 0.6. The experimental data is available for this range.} \\
  \todo{$S_{CD} -> S_{CH}$}
\end{figure}

@@

%depends on both, the
%binding affinity and the response of the head group on bound charge.
%Thus, the head group response to bound charge has to be correctly
%described in the model if the concept is used for quantitative studies of
%binding affinity. Hence, we first quantify the response of head group order
%parameters to the bound charge by using experimental data measured from
%mixtures of cationinc surfactants and PC lipids~\cite{scherer89}.
%
%Thus, the head group order parameter response to bound charge can be first
%quantified aginst experimental data before proceeding to binding affinity studies.
%Headgroup order parameter response to the amount of bound charge can
%be quantified by using ionic surfactants with a known valency~\cite{scherer89}.
%All surfactants locate in lipid bilayers due to their amphiphilic nature,
%thus the exact amount of bound charge in the membrane is known in these systems.
%The head group order parameter changes as a function of bound charges can be then
%explicitly determined from such systems.

The headgroup order parameter changes with increasing amount of 
cationic surfactant dihexadecyldimethylammonium bromide (C$_{12}$Cl$_{16}$$^+$N2C$_1$Br$^-$) 
is compared between experiments~\cite{scherer89} and simulations in
Fig. \ref{OrderParameterCHANGESsurf}. 
Approximately linear decrease is observed
in simulations and experiments with increasing amount of cationic
surfactants up to mole fractions of $\sim$50\%,
as expected from Eq. \ref{OPchangeEQ}.
The slope is, however, too steep in Lipid14 model indicating that 
the head group order parameters are too sensitive to a bound charge.
The ECC-lipids model gives a slope in very good agreement with experiments
for the $\alpha$ segment, while the slope is slightly
underestimated for the $\beta$ segment.
%This has to be taken into account when interpreting the
%head group order parameter response to CaCl$_2$ concentration in Lipid14 model.
%observed in  Ref. \citenum{catte16}, see also Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The present comparison of head group order parameter changes 
%suggests that the observed overestimated response of all models to 
%increasing CaCl$_2$ concentration in \cite{catte16} 
%can be explained at least in part by a high sensitivity of the head group order parameter response
%to the bound charge.
\todo{SAMULI: We could calculate the slopes from simulations, but I am not sure if
we would actually learn anything useful from this.}

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=16.0cm]{../Fig/OrdParChanges_NaCl_CaCl2.eps}
  \caption{\label{OrderParameterCHANGESnewMODELS}
    Changes of head group order parameters of POPC bilayer as a function of NaCl and CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure*}

@@

The binding affinities are quantified by using
the ion density profiles along membrane normal shown
in Fig.~\ref{fig:cacl-dens}. The density profiles show
larger Ca$^{2+}$ density peak in lipid headgroup region for
Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
The relative surface excess calculated from Eq.~\ref{surfexcess} gives 
$\Gamma_i^w = 0.07 nm^{-3}$ for the ECC-lipid model,
which is significantly smaller than $\Gamma_i^w = XX nm^{-3}$ for
Lipid14 with \AA{}qvist and $\Gamma_i^w = 0.3 nm^{-3}$ with Dang ions.
\todo{Surface excess for Lipid14 with Aqvist yet to be added.} \\
\todo{Surface excess units are not correct. Maybe the numbers needs to be checked also?} \\
\todo{JOE: Add error estimates to Gamma values}
Rough estimates for the free energy difference between bound and unbound cation
are given by
\begin{equation}
  \Delta G={\rm k_b T}\log(\frac{p_o}{p_i}),
\end{equation}
where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
in the maximum density in bilayer, respectively. The density profiles in
Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
in Lipid14 with \AA{}qvist.
